---
title: "Illustrate the distribution of PPV vs CNV-freq vs TP/FP before and after CNV-espresso"
[Initial] 10/22/2021
[Update]  02/03/2022

Renjie
output: html_notebook
---
```{r}
library(ggplot2)
library(scales) # for color palette
library("gridExtra") # for merging figures
library(ggpubr)
```

## 1. Import dataset. 
- The excel table contains XHMM raw calls on SPARK WES3 with overlapped WGS samples. We annotate WGS CNV calls identified by Canvas and Manta.
```{r}
cnv_df <- readxl::read_xlsx(path='./data/spark_wes3_wgs_xhmm_freq.xlsx')
rare_cnv_df <- cnv_df[which(cnv_df$`Num_Carriers(inGivenCohort)`<=6),] #569*1%=6
rare_cnv_df
```

## 2. Functions
```{r}
# [Figure 5 A,D,G] PPV vs CNV frequency
ppv_cnvfrequency <- function(data_df, cnv_frequency){
  num_sample <- length(unique(data_df$SAMPLE))
  selected_df <- data_df[which(data_df$`Num_Carriers(inGivenCohort)`/num_sample < cnv_frequency), ]
  TP_canvas <- nrow(selected_df[which(selected_df$TP_canvas!=''),])
  TP_union  <- nrow(selected_df[which(selected_df$TP_union !=''),])
  ppv_before_cnvEspresso <- TP_union/nrow(selected_df)
  
  selected_df_espresso  <- selected_df[which(selected_df$Status=='Positive'),]
  TP_canvas_espresso    <- nrow(selected_df_espresso[which(selected_df_espresso$TP_canvas!=''),])
  TP_union_espresso     <- nrow(selected_df_espresso[which(selected_df_espresso$TP_union !=''),])
  ppv_after_cnvEspresso <- TP_union_espresso/nrow(selected_df_espresso)
  return(c(ppv_before_cnvEspresso, ppv_after_cnvEspresso))
}

plot_ppv_cnvfreq <- function(cnv_df, legend_position){
  result_df <- data.frame(cnv_frequency=numeric(), ppv=numeric(), Tools=character(), stringsAsFactors = FALSE)

  for(num in 1:200){
    # here controls the cnv freq.
    cnv_frequency <- num/1000
    
    ppv <- ppv_cnvfrequency(cnv_df, cnv_frequency)
    print(paste0("CNV freq: ",ppv*100,"%"))
  
    result_df[nrow(result_df) + 1,] <- c(cnv_frequency, ppv[1], 'XHMM')
    result_df[nrow(result_df) + 1,] <- c(cnv_frequency, ppv[2], 'XHMM & CNV-espresso')
  }
  
  result_df$cnv_frequency <- as.numeric(result_df$cnv_frequency)
  result_df$ppv <- as.numeric(result_df$ppv)
  
  plot_ppv_cnvfreq <- ggplot(result_df, aes(x=cnv_frequency, y=ppv, colour=Tools)) +
                      geom_point(alpha=1) + 
                      #stat_smooth(method=lm) + # Smooth the dots
                      #scale_color_hue(direction = -1) + # reverse the color 
                      theme(text = element_text(size=18)) + 
                      ylab("Precision") + xlab("CNV frequency") + ylim(0.3,0.8) +
                      theme(legend.position=legend_position,
                            legend.background = element_rect(fill = "white", color = "black"), 
                            legend.title = element_blank(),
                            legend.text = element_text(size=10)) +
                      scale_color_manual(values = c("XHMM"=hue_pal()(3)[3], "XHMM & CNV-espresso"=hue_pal()(3)[1]))
  
  return(plot_ppv_cnvfreq)
}

# [Figure 5 B,E,F] PPV vs Number of targets
ppv_targets <- function(data_df, num_target){
  selected_df <- data_df[which(data_df$NUM_TARG >= num_target), ]
  TP_canvas   <- nrow(selected_df[which(selected_df$TP_canvas!=''), ])
  TP_union    <- nrow(selected_df[which(selected_df$TP_union !=''), ])
  ppv_before_cnvEspresso <- TP_union/nrow(selected_df) # we take TP_union for now, but we can also take step back.
  
  selected_df_espresso  <- selected_df[which(selected_df$Status=='Positive'), ]
  TP_canvas_espresso    <- nrow(selected_df_espresso[which(selected_df_espresso$TP_canvas!=''), ])
  TP_union_espresso     <- nrow(selected_df_espresso[which(selected_df_espresso$TP_union !=''), ])
  ppv_after_cnvEspresso <- TP_union_espresso/nrow(selected_df_espresso) # we take TP_union for now, but we can also take step back.
  
  return(c(ppv_before_cnvEspresso, ppv_after_cnvEspresso))
}

plot_ppv_numTargets <- function(rare_cnv_df, legend_position){
  result_df <- data.frame(num_target=numeric(), ppv=numeric(), Tools=character(), stringsAsFactors = FALSE)

  for(num_target in 1:20){
    ppv_targets(rare_cnv_df, num_target)
    #print(ppv_targets(rare_cnv_df, num_target))
    result_df[nrow(result_df) + 1,] <- c(num_target, ppv_targets(rare_cnv_df, num_target)[1], 'XHMM')
    result_df[nrow(result_df) + 1,] <- c(num_target, ppv_targets(rare_cnv_df, num_target)[2], 'XHMM & CNV-espresso')
  }
  
  result_df$num_target <- as.numeric(result_df$num_target)
  result_df$ppv <- as.numeric(result_df$ppv)
  
  plot_ppv_numTargets <- ggplot(result_df, aes(x=num_target, y=ppv, colour=Tools)) +
                          geom_line() +
                          geom_point(alpha=10) +
                          theme(text = element_text(size=18)) +
                          ylab("Precision") + xlab("Minimum number of targets") + ylim(0.3,0.8) +
                          theme(legend.position=legend_position,
                                legend.background = element_rect(fill = "white", color = "black"), 
                                legend.title = element_blank(),
                                legend.text = element_text(size=10)) +
                          scale_color_manual(values = c("XHMM"=hue_pal()(3)[3], "XHMM & CNV-espresso"=hue_pal()(3)[1]))
  return(plot_ppv_numTargets)
}


# [Figure 5 C,F,I] A bar plot to show the reduction performance between and after CNV-espresso
plot_reduction_performance <- function(TP, TP_espresso, TN, TN_espresso, legend_position){
  summary_data <- data.frame(cnv_status=rep(c("True  Postive", "False Postive"), each=2),
                     Tools=rep(c("XHMM", "XHMM & \nCNV-espresso"),2),
                     num_cnv=c(TP, TP_espresso, TN, TN_espresso))

  PLOT <- ggplot(summary_data, aes(x=Tools, y=num_cnv, fill=cnv_status)) + 
    geom_bar(stat="identity", width=0.7, position=position_dodge(width=0.7)) +
    geom_text(aes(label=num_cnv), vjust=0, size=6, position = position_dodge(width = 0.7)) +
    theme(text = element_text(size=18)) + ylab("Number of predictions") +  ylim(0, 2000) + 
    theme(axis.title.x=element_blank(), axis.text.x = element_text(color="black", size=18)) +
    theme(legend.position = legend_position, 
          legend.background = element_rect(fill = "white", color = "black"), 
          legend.title = element_blank(),
          legend.text = element_text(size=10)) +
    scale_fill_manual(values=c(hue_pal()(3)[3], hue_pal()(3)[1]))
  return(PLOT)
}
```

##  3. Analysis and Plots
### 3.1 Plots for the performance of initial rare CNVs before and after in silico validation by CNV-Espresso
- Not use this part in the manuscript, just for a test
```{r}
hist(rare_cnv_df$NUM_TARG,xlim=c(1,50),breaks=1000)
hist(cnv_df$CNV_freq,breaks=80)
```
```{r fig.height=3, fig.width=9}
plot_ppv_numTargets_unionAsTP <- plot_ppv_numTargets(rare_cnv_df)
plot_ppv_vs_cnvFreq_all <- plot_ppv_cnvfreq(cnv_df)

combo <- grid.arrange(plot_ppv_numTargets_unionAsTP, plot_ppv_vs_cnvFreq_all, fig_redu_canvas,  nrow=1, ncol =3)
#ggsave("./plots/combo.pdf", combo, width=15, height=5, unit="in")
combo
```

### 3.2 Plots for rare CNVs and **SQ > 60**
- Use this part in the manuscript
```{r}
## SQ >= 0 or 30
all_cnv_sq0 <- cnv_df[which(cnv_df$Q_SOME >= 0), ]
plot_ppv_freq_sq0 <- plot_ppv_cnvfreq(all_cnv_sq0, legend_position = c(0.68,0.85))

rare_cnv_sq0 <- rare_cnv_df[which(rare_cnv_df$Q_SOME >= 0), ]
plot_ppv_numTargets_unionAsTP_sq0 <- plot_ppv_numTargets(rare_cnv_sq0, legend_position = c(0.7,0.15))

## SQ >=60
all_cnv_sq60 <- cnv_df[which(cnv_df$Q_SOME >=60), legend_position = 'none' ]
plot_ppv_freq_sq60 <- plot_ppv_cnvfreq(all_cnv_sq60, legend_position = 'none')

rare_cnv_sq60 <- rare_cnv_df[which(rare_cnv_df$Q_SOME >=60), legend_position = 'none']
plot_ppv_numTargets_unionAsTP_sq60 <- plot_ppv_numTargets(rare_cnv_sq60, legend_position = 'none')

## SQ >=90
all_cnv_sq90 <- cnv_df[which(cnv_df$Q_SOME >=90), legend_position = 'none']
plot_ppv_freq_sq90 <- plot_ppv_cnvfreq(all_cnv_sq90, legend_position = 'none')

rare_cnv_sq90 <- rare_cnv_df[which(rare_cnv_df$Q_SOME >=90),  legend_position = 'none']
plot_ppv_numTargets_unionAsTP_sq90 <- plot_ppv_numTargets(rare_cnv_sq90,  legend_position = 'none')
```

### 3.3 Plots for rare CNVs and excluding SD regions
- Not use this part in the manuscript, just for a check
```{r}
rare_cnv_exlSD <- rare_cnv_df[which(rare_cnv_df$SD_region=='-' & rare_cnv_df$Q_SOME >=60),] #rare_cnv_df$`special_regions(clamms)`=='-' & 
plot_ppv_numTargets_unionAsTP_exlSD <- plot_ppv_numTargets(rare_cnv_exlSD)
plot_ppv_numTargets_unionAsTP_exlSD

cnv_sd_sq60 <- cnv_df[which(cnv_df$SD_region=='-' & cnv_df$Q_SOME >=60), ]
plot_ppv_freq_sd_sq60 <- plot_ppv_cnvfreq(cnv_sd_sq60)
plot_ppv_freq_sd_sq60

grid.arrange(plot_ppv_numTargets_unionAsTP_exlSD, plot_ppv_freq_sd_sq60, nrow=1, ncol = 2)

```

## 3.4 The performance for reducting TP and FP (reflect recall indirectly)
### 3.4.1 Canvas as GSD
- Not use this part in the manuscript
```{r}
rare_cnv_df <- cnv_df[which(cnv_df$`Num_Carriers(inGivenCohort)`<=6),]
TP_canvas <- nrow(rare_cnv_df[which(rare_cnv_df$TP_canvas!=""),])
TN_canvas <- nrow(rare_cnv_df[which(rare_cnv_df$TN_canvas!=""),])

rare_cnv_espresso_df <- rare_cnv_df[which(rare_cnv_df$Status=="Positive"),]
TP_canvas_espresso <- nrow(rare_cnv_espresso_df[which(rare_cnv_espresso_df$TP_canvas!=""),])
TN_canvas_espresso <- nrow(rare_cnv_espresso_df[which(rare_cnv_espresso_df$TN_canvas!=""),])

fig_redu_canvas <- plot_reduction_performance(TP_canvas, TP_canvas_espresso, TN_canvas, TN_canvas_espresso)
fig_redu_canvas
ggsave("./plots/cnvEspresso_reduction_performance_canvas.pdf", fig_redu_canvas, width=7, height=5, unit="in")
fig_redu_canvas
```

### 3.4.2 Using the intersection of canvas and Manta as GSD
- Not use this part in the manuscript
```{r}

TP_both <- nrow(rare_cnv_df[which(rare_cnv_df$TP_intersection!=""),])
TN_both <- nrow(rare_cnv_df[which(rare_cnv_df$TN_both!=""),])

rare_cnv_espresso_df <- rare_cnv_df[which(rare_cnv_df$Status=="Positive"),]
TP_both_espresso <- nrow(rare_cnv_espresso_df[which(rare_cnv_espresso_df$TP_intersection!=""),])
TN_both_espresso <- nrow(rare_cnv_espresso_df[which(rare_cnv_espresso_df$TN_both!=""),])

fig_redu_intersection <- plot_reduction_performance(TP_both, TP_both_espresso, TN_both, TN_both_espresso)
fig_redu_intersection
ggsave("./cnvEspresso_reduction_performance_canvas_manta.pdf", fig_redu_intersection, width=7, height=5, unit="in")
fig_redu_intersection
```

### 3.4.3 [Use this one] Using the union of canvas and Manta as GSD
```{r}
# SQ>=0
TP_union_sq0 <- nrow(rare_cnv_df[which(rare_cnv_df$TP_union!=""),])
TN_union_sq0 <- nrow(rare_cnv_df[which(rare_cnv_df$TN_both!=""),])

rare_cnv_espresso_df  <- rare_cnv_df[which(rare_cnv_df$Status=="Positive"),]
TP_union_espresso_sq0 <- nrow(rare_cnv_espresso_df[which(rare_cnv_espresso_df$TP_union!=""),])
TN_union_espresso_sq0 <- nrow(rare_cnv_espresso_df[which(rare_cnv_espresso_df$TN_both!=""),])

fig_redu_union_sq0 <- plot_reduction_performance(TP_union_sq0, TP_union_espresso_sq0, 
                                                 TN_union_sq0, TN_union_espresso_sq0, legend_position=c(0.75,0.87))
fig_redu_union_sq0

# SQ>=60
TP_union_sq60 <- nrow(rare_cnv_df[which(rare_cnv_df$TP_union!="" & rare_cnv_df$Q_SOME >=60 ),])
TN_union_sq60 <- nrow(rare_cnv_df[which(rare_cnv_df$TN_both!="" & rare_cnv_df$Q_SOME >=60),])

rare_cnv_espresso_df_sq60 <- rare_cnv_df[which(rare_cnv_df$Status=="Positive" & rare_cnv_df$Q_SOME >=60 ),]
TP_union_espresso_sq60 <- nrow(rare_cnv_espresso_df_sq60[which(rare_cnv_espresso_df_sq60$TP_union!=""),])
TN_union_espresso_sq60 <- nrow(rare_cnv_espresso_df_sq60[which(rare_cnv_espresso_df_sq60$TN_both!=""),])

fig_redu_union_sq60 <- plot_reduction_performance(TP_union_sq60, TP_union_espresso_sq60, 
                                                  TN_union_sq60, TN_union_espresso_sq60, legend_position='none')
fig_redu_union_sq60

# SQ>=90
TP_union_sq90 <- nrow(rare_cnv_df[which(rare_cnv_df$TP_union!="" & rare_cnv_df$Q_SOME >=90 ),])
TN_union_sq90 <- nrow(rare_cnv_df[which(rare_cnv_df$TN_both!="" & rare_cnv_df$Q_SOME >=90),])

rare_cnv_espresso_df_sq90 <- rare_cnv_df[which(rare_cnv_df$Status=="Positive" & rare_cnv_df$Q_SOME >=90 ),]
TP_union_espresso_sq90 <- nrow(rare_cnv_espresso_df_sq90[which(rare_cnv_espresso_df_sq90$TP_union!=""),])
TN_union_espresso_sq90 <- nrow(rare_cnv_espresso_df_sq90[which(rare_cnv_espresso_df_sq90$TN_both!=""),])

fig_redu_union_sq90 <- plot_reduction_performance(TP_union_sq90, TP_union_espresso_sq90, 
                                                  TN_union_sq90, TN_union_espresso_sq90, legend_position='none')
fig_redu_union_sq90
#ggsave("./plots/cnvEspresso_reduction_performance_union.pdf", fig_redu_union, width=7, height=5, unit="in")
```
## group sub-figures
```{r}
plot_ppv_freq_sq0
plot_ppv_freq_sq60
plot_ppv_freq_sq90
plot_ppv_numTargets_unionAsTP_sq0
plot_ppv_numTargets_unionAsTP_sq60
plot_ppv_numTargets_unionAsTP_sq90
fig_redu_union_sq0
fig_redu_union_sq60
fig_redu_union_sq90
combo <- grid.arrange(plot_ppv_freq_sq0,  plot_ppv_numTargets_unionAsTP_sq0,  fig_redu_union_sq0, 
                      plot_ppv_freq_sq60, plot_ppv_numTargets_unionAsTP_sq60, fig_redu_union_sq60, 
                      plot_ppv_freq_sq90, plot_ppv_numTargets_unionAsTP_sq90, fig_redu_union_sq90, 
                      nrow=3, ncol =3)
combo
ggsave("./plots/Figure5.pdf", combo, width=14, height=12, unit="in")
```

## 3.5 What's the CNV size distribution overlapped w/ calls from Canvas and Manta
```{r}
cnv_df <- readxl::read_xlsx(path='./data/spark_wes3_wgs_xhmm.xlsx')

unique(cnv_df$TP_union)
cnv_df_ov_manta  <- cnv_df[which(cnv_df$TP_union == "TP_manta"  | cnv_df$TP_union == "TP_canvas_manta"),]
cnv_df_ov_canvas <- cnv_df[which(cnv_df$TP_union == "TP_canvas" | cnv_df$TP_union == "TP_canvas_manta"),]

dim(cnv_df_ov_canvas)

plot_df1 <- cbind(cnv_df_ov_canvas, Tools="Canvas")
plot_df2 <- cbind(cnv_df_ov_manta, Tools="Manta")
plot_df  <- rbind(plot_df1, plot_df2)

ggplot(plot_df, aes(NUM_TARG, fill=Tools, colour=Tools)) +
  geom_density(alpha = 0.3) +
  xlim(0, 50) +
  theme(text = element_text(size=18)) + 
  scale_fill_manual(values=c(hue_pal()(3)[3], hue_pal()(3)[1])) +
  scale_color_manual(values=c(hue_pal()(3)[3], hue_pal()(3)[1]))
```

```{r}
cnv_df
# for rare CNVs
plot_rare_df <- plot_df[which(plot_df$`Num_Carriers(inGivenCohort)`<=6),]

ggplot(plot_rare_df, aes(NUM_TARG, fill=Tools, colour=Tools)) +
  geom_density(alpha = 0.3) +
  xlim(0, 50) +
  theme(text = element_text(size=18)) + 
  scale_fill_manual(values=c(hue_pal()(3)[3], hue_pal()(3)[1])) +
  scale_color_manual(values=c(hue_pal()(3)[3], hue_pal()(3)[1]))
```

